Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-18 manu-draft.qmd
update prior and sensitivity rationale from RN suggestions

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$. Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values. The standard deviation of $R$ was 170 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $aPAR$ = $R$. The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values. Together, the default priors for $a$, $R$, and $b$ are:

$$
a \sim N(4, 2.25)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
R \sim N(300, 170)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.126)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default standard deviations to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

>>>>

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for BASE, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$. Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004. The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$. The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. For the prior standad deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14. However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal. Together, the default priors for $a$, $R$, and $b$ are:

$$
a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.
